Operational Feature Selection in Gaussian Mixture Models

### 2016-09-27 paper.tex
Section II.A et B
+  une ref en plus.

\documentclass[journal,peerreview,onecolumn]{IEEEtran}
% \documentclass[journal]{IEEEtran}

\usepackage{cite}
\usepackage[pdftex]{graphicx}
\graphicspath{{./Fig/}}
\DeclareGraphicsExtensions{.pdf,.jpg,.png}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{array}
\usepackage[caption=false,font=footnotesize]{subfig}
\usepackage{fixltx2e}
\usepackage{url}
\usepackage{listings}
\usepackage{multirow}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{tabularx}
\usepackage{empheq}

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
This article presents a forward feature selection algorithm based on Gaussian mixture model (GMM) classifiers. The algorithm selects iteratively features that maximize a criterion function which can be either a classification rate or a measure of divergence. Several variations of this algorithm are explored by changing the criterion function and also by testing a floating forward variation allowing backward step to discard already selected features.

@@

\section{Gaussian Mixture Models in high dimension space}
\label{sec:gmm-hd}

In the remaining of the paper, the following notations are used. $\mathcal{S} = \{\mathbf{x}_i,y_i\}_{i=1}^{n}$ denotes the training set where $\mathbf{x}_i \in \mathbb{R}^d$ is the vector of features of the $i^{th}$ sample, $y_i = 1,...,C$ the associated label, C the total number of classes, $n$ the number of samples and $n_c$ the number of samples of class $c$.

@@

    The hypothesis of mixture models is that a given sample is the realization of a random vector which distribution is a mixture (convex combination) of several class conditioned distribution:
    \begin{equation}
        p(\mathbf{x}) = \sum_{c=1}^{C} \pi_c f_c(\mathbf{x}|\theta)
    \end{equation}
    where $\pi_c$ is the prior i.e. the proportion of class $c$ and $f_c$ a probability density function parametrized by $\theta$.

    The Gaussian mixture model (GMM) assumes that each $f_c$ is , conditionally to $c$, a Gaussian distribution of parameters $\boldsymbol{\mu}_c$ and $\boldsymbol{\Sigma}_c$ and so $f_c(\mathbf{x}|\theta)$ can be written as
    \begin{equation*}
        f_c(\mathbf{x}|\theta) = \frac{1}{(2\pi)^{\frac{d}{2}} |\boldsymbol{\Sigma}_c|^{\frac{1}{2}}} \exp \left( -\frac{1}{2} (\mathbf{x} - \boldsymbol{\mu}_c)^t \boldsymbol{\Sigma}_c^{-1} (\mathbf{x} - \boldsymbol{\mu}_c) \right).
    \end{equation*}

    In supervised learning the class parameters $\boldsymbol{\mu}_c$, $\boldsymbol{\Sigma}_c$ and $\pi_c$ are usually estimated through the conventional unbiased empirical estimators:
    \begin{align}
        \hat{\pi}_c &= \frac{n_c}{n},\\
        \hat{\boldsymbol{\mu}}_c &= \frac{1}{n_c} \sum_{\{i|y_i = c\}} \mathbf{x}_i ,\\
        \hat{\boldsymbol{\Sigma}}_c &= \frac{1}{(n_c - 1)} \sum_{\{i|y_i = c\}} (\mathbf{x}_i - \boldsymbol{\mu}_c) (\boldsymbol{}x_i - \boldsymbol{\mu}_c)^t.
    \end{align}

    To predict the class of a new sample, the maximum a posteriori (MAP) rule is used and thus, using Bayes' law, the decision rule is
    \begin{equation*}
        \mathbf{x} \text{ belongs to } c \Leftrightarrow c = \text{arg} \max_{c \in C} p(c) p(\mathbf{x}|c).
    \end{equation*}
    Identifying $p(c)$ as $\pi_c$ and $p(x|c)$ as $f_c(x|\theta)$ and by taking the log, a simplified decision rule is obtained
    \begin{align}
        Q_c(\mathbf{x})
        &= 2 \log \left( p(c) p(\mathbf{x}|c) \right) \nonumber \\
        &= - (\mathbf{x} - \boldsymbol{\mu}_c)^t \boldsymbol{\Sigma}_c^{-1} (\mathbf{x} - \boldsymbol{\mu}_c) \nonumber \\
        &~~-\log (|\boldsymbol{\Sigma}_c|) + 2 \log (\pi_c) - d \log (2\pi).
        \label{eq:decision}
    \end{align}

    It is important to notice that the inverse and the determinant of the covariance matrix are key elements of the decision function. The estimations of these elements suffer from the curse of dimensionality\cite{bouveyron2014model}. More precisely, the number of parameters to estimate increases quadratically relatively to the number of features as illustrated by Figure~\ref{fig:nb-param}. The minimum number of samples requires to estimate the model is equal to the number of parameters and it can be an issue in some case. For example, hyperspectral data is a case of high dimensional samples and, in general, very few labeled samples are available because of the difficulty and the cost to collect ground-truth.

    \begin{figure}[!t]
        \centering
        \begin{tikzpicture}
            \begin{axis}[xmin=0,xmax=500,ymin=0,width=0.5*\columnwidth,grid,axis x line=left ,axis y line=left, tick align=outside,xlabel = Dimension,ylabel =Number of parameters]
                \addplot+[very thick,mark=none,smooth,domain=0:500,samples=500] (\x,{\x*(\x+3)/2+1});
            \end{axis}
        \end{tikzpicture}
        \caption{Number of parameters by class in function of dimension.\label{fig:nb-param}}
    \end{figure}

    A lack of samples induces badly-conditioned covariance matrices and so unstable inversion and computation of the determinant. There are two major solutions to this problem. First option is to use a regularization method to stabilize the inversion of the covariance matrices\cite{hoerl1970ridge}. Second option is to use a features extraction/selection method in order to reduce the dimensionality of the samples. In this study, the latter option is explored and a feature selection method named sequential forward features selection is presented in the next Section.

    \subsection{Features selection}
    \label{sec:selection}

    The objective of feature selection is to retrieve the most compact representation of the data with minimum loss of information. In deleting the redundant information, feature selection allows to reduce dimension and thus avoid statistical issue, increase algorithm speed and limit storage requirements. Moreover, it adds several beneficial side effects. It improves data understanding by identify where is the most relevant information and it is then possible to save resources when organizing a new data acquisition.

@@

% Note that the IEEE does not put floats in the very first column
% - or typically anywhere on the first page for that matter. Also,
% in-text middle ("here") positioning is typically not used, but it
% is allowed and encouraged for Computer Society conferences (but
% not Computer Society journals). Most IEEE journals/conferences use
% top floats exclusively.
% Note that, LaTeX2e, unlike IEEE journals/conferences, places
% footnotes above bottom floats. This can be corrected via the
% \fnbelowfloat command of the stfloats package.

>>>>

\documentclass[journal,peerreview,onecolumn]{IEEEtran}
% \documentclass[journal]{IEEEtran}
\usepackage{etex}
\usepackage{cite}
\usepackage[pdftex]{graphicx}
\graphicspath{{./Fig/}}
\DeclareGraphicsExtensions{.pdf,.jpg,.png}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{array}
\usepackage[caption=false,font=footnotesize]{subfig}
\usepackage{fixltx2e}
\usepackage{url}
\usepackage{listings}
\usepackage{multirow}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{tabularx}
\usepackage{empheq}
\usepackage{hyperref}

@@

\tableofcontents
\clearpage
% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
This article presents a forward feature selection algorithm based on Gaussian mixture model (GMM) classifiers. The algorithm selects iteratively features that maximize a criterion function which can be either a classification rate or a measure of divergence. Several variations of this algorithm are explored by changing the criterion function and also by testing a floating forward variation allowing backward step to discard already selected features.

@@

\section{Gaussian Mixture Models in high dimensional space}
\label{sec:gmm-hd}

The    following    notations    are   used    in    the    remaining.
$\mathcal{S}  = \{\mathbf{x}_i,y_i\}_{i=1}^{n}$  denotes the  training
set where $\mathbf{x}_i \in \mathbb{R}^d$ is the vector of features of
the $i^{th}$ sample, $y_i = 1,...,C$ the associated label, C the total
number of classes,  $n$ the number of samples and  $n_c$ the number of
samples of class $c$.

@@

For mixture models, it is assumed  that a given sample $\mathbf{x}$ is
the realization  of a  random vector which  distribution is  a mixture
(convex     combination)     of      several     class     conditioned
distribution~\cite{Fraley00model-basedclustering}:
    \begin{equation}
        p(\mathbf{x}) = \sum_{c=1}^{C} \pi_c f_c(\mathbf{x}|\theta)
    \end{equation}
    where $\pi_c$ is  the prior, \emph{i.e.}, the  proportion of class
    $c$  and  $f_c$  a  parametric density  function  parametrized  by
    $\theta$.

    Among the possible parametric model,  the Gaussian one is the most
    used~\cite{bouveyron2014model}.   It assumes  that each  $f_c$ is,
    conditionally  to  $c$,  a  Gaussian  distribution  of  parameters
    $\boldsymbol{\mu}_c$    and    $\boldsymbol{\Sigma}_c$:
    \begin{equation*}
        f_c(\mathbf{x}|\boldsymbol{\mu}_c, \boldsymbol{\Sigma}_c) = \frac{1}{(2\pi)^{\frac{d}{2}} |\boldsymbol{\Sigma}_c|^{\frac{1}{2}}} \exp \left( -\frac{1}{2} (\mathbf{x} - \boldsymbol{\mu}_c)^t \boldsymbol{\Sigma}_c^{-1} (\mathbf{x} - \boldsymbol{\mu}_c) \right).
    \end{equation*}

    It  is  referred  to  as  Gaussian  mixture  model  (GMM).   In  a
    supervised    learning    framework,    the    class    parameters
    $\boldsymbol{\mu}_c$,   $\boldsymbol{\Sigma}_c$   and  the   prior
    $\pi_c$ are  usually estimated  through the  conventional unbiased
    empirical estimators:
    \begin{align}
        \hat{\pi}_c &= \frac{n_c}{n},\\
        \hat{\boldsymbol{\mu}}_c &= \frac{1}{n_c} \sum_{\{i|y_i = c\}} \mathbf{x}_i ,\\
        \hat{\boldsymbol{\Sigma}}_c &= \frac{1}{(n_c - 1)} \sum_{\{i|y_i = c\}} (\mathbf{x}_i - \boldsymbol{\mu}_c) (\boldsymbol{}x_i - \boldsymbol{\mu}_c)^t.
    \end{align}

    To predict the  class of a new unseen sample,  the maximum \emph{a
      posteriori}  rule  is  used:
    \begin{equation*}
        \mathbf{x} \text{ belongs to } c \Leftrightarrow c = \text{arg} \max_{c \in C} p(c) p(\mathbf{x}|c).
    \end{equation*}
    Under the GMM,  and identifying $p(c)$ as $\pi_c$  and $p(x|c)$ as
    $f_c(x|\theta)$ and by taking the log, a decision rule is obtained
    \begin{eqnarray}\label{eq:decision}
      Q_c(\mathbf{x}) &=& 2 \log \left( p(c) p(\mathbf{x}|c) \right) \nonumber \\
                      &=& - (\mathbf{x} - \boldsymbol{\mu}_c)^t \boldsymbol{\Sigma}_c^{-1} (\mathbf{x} - \boldsymbol{\mu}_c) \nonumber \\
                      & &-\log (|\boldsymbol{\Sigma}_c|) + 2 \log (\pi_c) - d \log (2\pi).
    \end{eqnarray}

    \subsection{Curse of dimensionality in GMM}
    \label{sec:curse:gmm}

    The computation of~\ref{eq:decision} requires the inversion of the
    covariance matrix and and the  computation of the logarithm of the
    determinant.  The estimations of these terms suffer from the curse
    of  dimensionality\cite{bouveyron2014model}, because  in practice,
    the  number of  parameters  $\eta_c$ to  estimate  for each  class
    increases quadratically with respect to the number of features, as
    illustrated in Figure~\ref{fig:nb-param}.  Hence, if the number of
    observation  $n_c$ is  small  compared to  the  number of  samples
    $n_c$, the  estimated covariance  matrix is badly  conditioned and
    thus the computation  of its inverse and its  determinant would be
    unstable.  The  worst situation  is  $n_c<\eta_c$  which leads  to
    singular covariance matrix.

    Unfortunately, this situation happens regularly in remote sensing.
    For  instance  in  hyperspectral image  classification,  very  few
    labeled samples  are usually  available because of  the difficulty
    and the cost to collect ground-truth.

    \begin{figure}[!t] % TODO : log plot et donner la formule dans la légende 
        \centering
        \begin{tikzpicture}
          \begin{semilogyaxis}[xmin=0,xmax=200,ymin=0,width=0.5*\columnwidth,grid,axis x line=left ,axis y line=left, tick align=outside,xlabel = $d$,ylabel = $\eta_c$]
            \addplot+[very thick,mark=none,smooth,domain=0:200,samples=200] (\x,{\x*(\x+3)/2+1});
            \end{semilogyaxis}
        \end{tikzpicture}
        \caption{Number of parameters $\eta_c$ by class in function of dimension $d$: $\eta_c=d(d+3)/2+1$.\label{fig:nb-param}}
    \end{figure}

    %% TODO: il faut reprendre cette partie en détaillant plus
    %% Il y a deux approches: une qui regularise le probleme inverse,
    %% ridge est une solution, mais il y en a d'autres. Voir le
    %% paragraphe qui est consacré à cela dans la thèse d'Anthony ->
    %% elle arrive pas mail. Ensuite, il y a la réduction de dimension
    %% données des refs (voir intro du papier JSTARS) et dans la
    %% réduction  de dimension, il  y a  la sélection de  features. La
    %% conclusion doit s'enchainer avec la partie suivante.
    There are two major  solutions to this problem. First
    option  is  to  use  a  regularization  method  to  stabilize  the
    inversion of the  covariance matrices\cite{hoerl1970ridge}. Second
    option is to  use a features extraction/selection  method in order
    to reduce  the dimensionality of  the samples. In this  study, the
    latter option  is explored  and a  feature selection  method named
    sequential  forward features  selection is  presented in  the next
    Section.

    \section{Features selection}
    \label{sec:selection}

    The objective of feature selection is to retrieve the most compact
    representation of  the data with  minimum loss of  information. By
    deleting the  redundant information,  feature selection  allows to
    reduce  dimension  and  thus  avoid  statistical  issue,  increase
    algorithm speed and limit  storage requirements. Moreover, it adds
    several beneficial side effects. It improves data understanding by
    identify where  is the  most relevant information  and it  is then
    possible to save resources when organizing a new data
    acquisition.% TODO: add ref. et example.

@@

% Note that the IEEE does not put floats in the very first column
% - or typically anywhere on the first page for that matter. Also,
% in-text middle ("here") positioning is typically not used, but it
% is allowed and encouraged for Computer Society conferences (but
% not Computer Society journals). Most IEEE journals/conferences use
% top floats exclusively.
% Note that, LaTeX2e, unlike IEEE journals/conferences, places
% footnotes above bottom floats. This can be corrected via the
% \fnbelowfloat command of the stfloats package.
%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
